---
title: "Chapter 5. DIY: Entity Resolution"
output:
  pdf_document: default
  html_document: default
---

### How do I auto-populate text and stences pro forma?
  
In 2014, an automated script was the first produced by the L.A. Times was the first to [report on an earthquake around Los Angeles](http://www.slate.com/blogs/future_tense/2014/03/17/quakebot_los_angeles_times_robot_journalist_writes_article_on_la_earthquake.html). Everytime the United States Geological Survey issues an earthquake over a certain severity level, the robot is able to pick out key pieces of information and populate an article _pro forma_. In the wild, many well-structured data sources that are issued as alerts are well-suited for automation.  What if we have the following statement.  What if specific information could be auto-populated? 

\vspace{12pt} 
```{r}
  original <- "Maria was a Category 5 with max wind speeds of 175 mph."
```
\vspace{12pt} 

We need to first find key items and replace with unique placeholders. The unique placeholders can be used as a standardized search. Notice that each modification to the original string `original` is assigned to the new object `warn.line` in order to avoid overwriting the gold copy.

\vspace{12pt} 
```{r}
#Replace Category 5 with <hurricane_level>
  warn.line <- gsub("Category \\d", "Category <hurricane_level>", original)
  
#Search for Maria as a five-letter word with <name>
  warn.line <- gsub("\\b[[:alpha:]]{5}\\b", "<name>", warn.line)
  
#Replace 175 with <wind_speed>
  (warn.line <- gsub("\\d{3}", "<wind_speed>", warn.line))
  
```
\vspace{12pt} 

As information becomes available, this sentence can be populated with event-specific information. Below is a list of a few hurricanes that occurred in the 2017 season.

\vspace{12pt}   
```{r}
  hurricanes <- rbind(data.frame(storm.name = "Irma", speed = 185, level = 5),
            data.frame(storm.name = "Jose", speed = 155, level = 4),
            data.frame(storm.name = "Katia", speed = 105, level = 2),
            data.frame(storm.name = "Harvey", speed = 130, level = 4))
```
\vspace{12pt} 

To auto-populate the sentence, we can write a simple function that accepts key inputs, then outputs a storm summary. Note that the function includes descriptive text to help future users understand how to use the code. 

\vspace{12pt} 
```{r}
  warningText <- function(model.sentence, storm.name, level, speed){
    #
    # Converts key hurricane information into summary text
    #
    # Args:
    #   model.sentence = model sentence with markers 
    #   storm.name = string containing storm name to replace <name>
    #   level = hurricane level to replace <hurricane_level>
    #   speed = maximum speed in mph to replace <wind_speed>
    #
    # Result:
    #   Storm summary populated with storm details
    #
    
    statement <- gsub("<hurricane_level>", level, model.sentence)
    statement <- gsub("<wind_speed>", speed, statement)
    statement <- gsub("<name>", storm.name, statement)
    return(statement)
  }
```
\vspace{12pt} 

With the function, we can loop through each row of the data frame to generate a sentence describing each storm.

\vspace{12pt} 
```{r}
for(i in 1:nrow(hurricanes)){
  print(warningText(model.sentence = warn.line, 
                    storm.name = hurricanes$storm.name[i], 
                    level = hurricanes$level[i], 
                    speed = hurricanes$speed[i]))
}
  
```