--- 
title: "Data Science + Public Policy"
author: "Jeff Chen"
date: '`r Sys.Date()`'
output: pdf_document
description: Chapter 10
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
---

### How much do people earn based on their characteristics? 

This exercise can be viewed in two ways:

- the traditional econometric approach would focus on interpretation of the coefficients as marginal effects of various factors
- the data science approach would design a model validation framework to develop stable, reliable model that could be used to score future data


- Note that this data usually has sampling weights associated with each record due to how the data is collected. For convenience, the sampling weights are set aside.


```{r, eval = FALSE}
#ACS file is large -- this might a minute or two
library(digIt)
acs <- digIt("acs_wages")
```

#### Split Training Set

```{r}
train <- acs[acs$partition == "train", ]
test <- acs[acs$partition == "test", ]
```

#### Model 
```{r}
mod <- lm(log(wagp)~ wkhp + naics + schl + sex + esr + agep, data = train)
yhat <- predict.lm(mod, newdata = test)
```