--- 
title: "Data Science + Public Policy"
author: "Jeff Chen"
date: '`r Sys.Date()`'
output: pdf_document
description: Chapter 10
documentclass: book
link-citations: yes
bibliography:
- book.bib
- packages.bib
site: bookdown::bookdown_site
biblio-style: apalike
---

## Exercises

####  Prediction {-}
Using the Maryland Toll Road data, try the following: 

1. Run a regression for `FIPS = 24015` where $Y$ is $log(transactions)$ and $X$ are log(emp), log(bldgs), and  log(wti_eia).
2. Obtain the training and testing MAPEs for `FIPS = 24015`. How does this compare to the MAPE for `FIPS = 24015` from the example pooled model?

#### Calculating Distance {-}

1. Write a function `dist()` that calculates the distance between all records of a two variable data frame `df` and a given reference coordinate. The reference coordinate will be an index $i$ of a row in the data frame (e.g. calculate the distance between row $i$ and all other points).
2. Expand that distance function to accept one or more variables. Use the example data to test your function.
```{r}
  set.seed(150)
  data <- data.frame(x1 = rnorm(1000, 10, 5), 
                     x2 = rnorm(1000, 20, 35), 
                     x3 = rnorm(1000, 14, 1), 
                     x4 = rnorm(1000, 100, 200))
```
3. For the following values, write a function to retrieve the value of $y$ where $k = 1$ for each record $i$. 
4. Modify the function to handle $k = 2$.
    

#### Nearest Neighbors {-}

1. Write a function `dist()` that calculates the distance between all records of a two variable data frame `df` and a given reference coordinate. The reference coordinate will be an index `i` of a row in the data frame (e.g. calculate the distance between row `i` and all other points).
```{r}
  dist <- function(df, i){
    temp <- df[i,]
    dist <- sqrt(df[,1]^2 + df[,1]^2)
    return(dist)
  }
```

2. Expand that distance function to accept one or more variables. 
```{r}
#Modify the function
  dist <- function(df, i){
    temp <- df[i,]
    col <- ncol(df)
    dist <- 0
    for(k in 1:col){
      dist <- dist + (df[,k] - temp[,k])^2
    }
    return(sqrt(dist))
  }

#Test it
  set.seed(150)
  data <- data.frame(x1 = rnorm(1000, 10, 5), 
                     x2 = rnorm(1000, 20, 35), 
                     x3 = rnorm(1000, 14, 1), 
                     x4 = rnorm(1000, 100, 200))
  out <- dist(data, 1)
  head(out)
```

3. Write a function to retrieve the value of $y$ where $k = 1$ for each record $i$. 
4. Modify the function to handle $k = 2$.